---
title: "Using tidyr to wrangle data"
author: "Erica Bishop"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

library(tidyverse)
library(here)
library(janitor)
```

## Read in WB data files

```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..",""))

```

Now tidy up data

```{r}
#step one: get years into one column

#use single or double quotes around column names with numbers and spaces to make sure r knows what you're referring to

wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = `2001 [YR2001]`:`2020 [YR2020]`,
               names_to = "year",
               values_to = "indicator_value")

#note that year is a character class

```

Now use `separate()` to clean up the year column - split into year and year character

```{r}
#filter is for rows, select is for columns

wb_clean <- wb_indicators_long |> 
  separate(col = year, into = c("year", "year_chr"), sep = " ") |>
  select(-year_chr, -'Country Code', -'Series Code') |> 
  mutate(year = as.numeric(year)) |> 
  drop_na('Series Name') |> 
  pivot_wider(names_from = "Series Name", values_from = "indicator_value")

#Now the data is tidy!

#always look at the intermediate steps before deciding on your next step

#use pivot wider to widen variables trapped in 'Series Name' to be spread acorss multiple columns and get the values that populate the columns from indicator_value
#drop NA will drop an entire row with an NA value so BE CAREFUL!!

```

To see all of the unique values that exist in a series:

```{r}
unique(wb_clean$`Series Name`)
```

```{r}
#redname column headers so they aren't so unweildy
#with names(), be careful with order of things

names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")

#can't pipe straight into names, that function is one that isn't built to accept a piped in argument

```

Now create a subset and plot (filter for rows)

```{r}

wb_subset <- wb_clean |> 
  filter(country %in% c("Canada", "China", "Tajikistan", "Norway"))

ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt)) +
  geom_line(aes(color = country)) +
  facet_wrap(~country)
  

```

For example, if we wanted to get rid of some of the country class variables that are larger region groupings, etc., then use:

```{r}
wb_example <- wb_subset |> 
  filter(!country %in% exclude_this)

#this gets rid of c
```

## More data wrangling with filter

```{r}
ex_1 <- starwars |> 
  filter(height > 180)
```

```{r}
ex_2 <- starwars |> 
  filter(eye_color == "blue")
#check it:
unique(ex_2$eye_color)

```

```{r}

ex_3 <- starwars |> 
  filter(homeworld == "Naboo")

```

Now try filtering with multiple conditions:

```{r}
ex_4 <- starwars |> 
  filter(height > 180 & homeworld == "Tatooine")

```

```{r}
ex_5 <- starwars |> 
  filter(hair_color == "brown" & species == "Human")
```

## Filtering `or` `|` statements

```{r}
ex_6 <- starwars |> 
  filter(height > 180 | eye_color == "yellow")
```

create a subset that keeps rows if homeworld is Endor or species droid

```{r}
ex_7 <- starwars |> 
  filter(homeworld == "Endor" | species == "Droid")

```

ANother way to write an or statement referring to a single variable's values:

Ex. keeping any observations where eye color is blue or red or brown; instead of writing out each of the matches we are looking for with multiple or statements, use %in%

```{r}
ex_8 <- starwars |>
  filter(eye_color %in% c("blue", "brown", "red"))

#the in operator looks at each row and looks for any match within the vector, blue, or brown, or red

# the == is only for looking for a single exact match, NOT a vector.... otherwise:

#DON'T DO THIS!!::
ex_BAD <- starwars |> 
  filter(eye_color == c("blue", "brown", "red"))
#this is looking for this specific order, so row 1 looks for match to blue, row 2 looks for match to brown, row 3 looks for match to red, row 4 to blue, etc. etc.
  
```

```{r}

ex_11 <- starwars |> 
  filter(homeworld %in% c("Endor", "Tatooine", "Naboo") 
         & height < 150)


```

Using a negate filter

```{r}
ex_12 <- starwars |> 
  filter(species != "Human")

ex_13 <- starwars |> 
  filter(!species %in% c("Human", "Ewok", "Wookiee"))

```

## `select()` to select or exclude columns

Some examples using the cleaned worldbank dataset

```{r}
#select by name

ex_14 <- wb_clean |> 
  select(country, year, co2_emissions_kt)

#or refer to a range of values - select is a risky function to use becasue there is no relationship order to the columns, purely positional so can be changed. But can be useful to view a simplified dataset

ex_15 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct)

#exclude variables in a range

ex_16 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct, -access_electricity_pp)

#now pipe some things together

ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)

```

```{r}
#can also use select to rename and/or move things - below columns are re-ordered and renamed

ex_18 <- wb_clean |> 
  select(year_new = year,
         emissions = co2_emissions_kt,
         country_name = country)
```

```{r}
ex_19 <- wb_clean |> 
  select(year, country, water_stress) |> 
  filter(country %in% c("Albania", "Croatia", "Ecuador")) |> 
  drop_na()

#check to see if it worked

unique(ex_19$country)

head(ex_19$country)

```

Another way to rename columns with `rename()`

```{r}

ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)

#select makes you put in every column even the ones you don't want to rename, so using rename allows you to just call out the ones you do want to rename, and leaving the rest

ex_21 <- wb_clean |> 
  rename(place = country,
         electricity_access = access_electricity_pp)


```

## `mutate()` to update an existing column or add a new one

If you assign a NEW column name it will add a new column to the end of your data frame. If you assign an existing column name, it will overwrite that column.

Both can be useful depending on what you want to do.

**Adding a new column:** Example: converting units of CO2 emissions from kt to tons - 100 ktons / 1 ton

```{r}
#create a new column
ex_22 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt*1000) |> 
  relocate(co2_emissions_t, .after = co2_emissions_kt)

#make another new column with new values - this will help build to using casewhen()

ex_23 <- wb_clean |> 
  mutate(yay = "YAY NEW COLUMN")

#can also use to recast variables as new classes

ex_24 <- wb_clean |> 
  mutate(year = as.character(year))

ex_25 <- ex_24 |> 
  mutate(year = as.numeric(year))

#dplyr::relocate() to move order of columns - this doesn't really have an anlytical function, but can just visually make things feel better

ex_26 <- wb_clean |> 
  relocate(water_stress, .after = year)


```

## `group_by() |> summarize()`

This combo creates groups in dataset based on variables you specify - so then anything done after is applied to those groups. `summarize()` applies functions to EACH GROUP and returns them in a summary table.

If you run just group_by() you won't *see* anything, but then when you pipe to a next step or apply another function it will be applied by group

```{r}
ex_27 <- starwars |> 
  group_by(species) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))


```

```{r}
#create a case table to make this tidy

df <- tribble(
~species, ~length, ~number,
"lobster", 12, 4,
"lobster", 14, 6,
"lobster", 15, 2
)

df_case <- df |> 
  uncount(number)
```
 `uncount()` creates table that has a single row for each observation from a frequency or count table

`count()` does the opposite - counts observations and is a combination function of group_by, summarize, n() - so jut feed n() as the argument

For example:
```{r}
starwars |> count(homeworld)

#can give multiple conditions

starwars |> count(homeworld, species)

starwars |> count(species, homeworld)

```

Another practice example: 

```{r}

ex_28 <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile"),
         year > 2005) |> 
  select(country, year, co2_emissions_kt) |>
  mutate(co2_emissions_Mt = co2_emissions_kt / 1000)

total_emissions <- ex_28 |> 
  group_by(country) |> 
  summarize(total_co2_Mt = sum(co2_emissions_Mt, na.rm = TRUE))

```

Remember to check intermediates - max, min, names, unique, etc. in the console

